### 2021-09-19 plos_latex_template.tex
Rocío: Rewriting rule online

The rules written here are directed to people who are part of a stable meetings committee that oversees the site/location selection process, that coordinates with the local organizers of conferences, or even the local/virtual organizers who desire to make an inclusive conference starting at the planning stage.
These tips stem from the authors' experience of organizing useR! 2021, a virtual and global statistical computing conference for users and developers of the R programming language \cite{r_core_team_2021}. 
We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, truly inclusive and accessible experience for everyone. 
Here, we share the lessons learned within the past year of organizing this global useR!2021, summarized as 10 simple rules towards an inclusive conference.

@@

%This is specially important for networking spaces, where receiving ideas, direct feedback, and spaces to co-lead from a wide variety of community members can be the key to organize meaningful sessions for all the conferences attendees. This gave a sense of variety and an appeal of “at least one networking session for everyone”. 
%ast 17sep: i think we can talk here about planning neworking here, but also: the whole paragraph applies to team dynamics in general. Sara: maybe you can use this as we spoke relating team work here. And I think the working together part is not here but the following paragraph 
%sara: The last sentence in the paragraph sounds odd to me, but maybe I'm missing something. Andrea, I added a few sentences to the following paragraph regarding work dynamics

@@

Make sure that every selection committee --the committee looking for keynote speakers, the selection committee for abstracts, the prizes and award committees --are also diverse \cite{swartzScienceValueDiversity2019, wongBuildDiversityScience2020}, and ask them to be aware that everyone has implicit biases, to recognize them, and try to counteract them. 
An inclusive and diverse organizing team (\textbf{Rule \ref{rule_organizing_team}}) is already a great starting point to overcome this bias in other roles. The regional and local communities in your field are also good sources to tap into. 
For example, for useR! 2021, groups like AfricaR (africa-r.org), R-Ladies (rladies.org), MiR (mircommunity.com), Forwards (forwards.github.io), and LatinR (latin-r.com) were fundamental to reach people for the organizing team, potential presenters, to co-lead social events, and attendees; some of them even held spaces to help the members of their groups to prepare abstracts for submission.
When organizing an international academic conference, you may reach local associations, student organizations or other Early Career Researchers groups or communities of practice: ask for their requirements and barriers, and explicitly invite them to contribute to the conference. Since many conferences are held with the primary goal of displaying cutting-edge work led by senior scientists and academics, early career researchers are often disadvantaged. Diversity should spread among all speaking roles, and the conference should also help raise the profiles of early career researchers, fostering collaboration and building their skills.

Furthermore, consider bringing to the spotlight the people that have contributed to your field in more collaborative ways \cite{cheng2020x+}. You can, for instance, reframe the awards ceremony to acknowledge those who prepared accessible slides and presentations (see Rule \ref{rule_accessibility}), to acknowledge being mindful of inclusiveness. Community building, for instance, is challenging and usually unrewarded when compared to publications or software development \cite{acionWhyChooseCommunity2020}; give the people who deliberately took the time to work on their communities the recognition they deserve. Defy the stereotypical criteria for success by acknowledging these community practices. 
Finally, do not restrict people from marginalized groups or community-builders to talk or work only in issues related to Diversity, Inclusion, and Accessibility. Recognize their areas of expertise and respect their will regarding participation (or not) in community-building events.

@@

In-person interaction is priceless; however, going to conferences is expensive, and many people do not attend for lack of financial support. This inaccessibility %ast: maybe inaccessibility is not the word here because we want to focus on disabilities with this word
is particularly true for global conferences that usually take place in high-income countries, making it financially demanding for international participants and often impossible to attend due to immigration requirements \cite{arendDisparityConferenceRegistration2019,gewinWhatScientistsShould2019}. Online conferences are more inclusive: they do not require a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities. This means that online conferences have a greater reach, not only in terms of participants but in terms of the tutors and presenters that can participate \cite{atkinsonJournalMedicine20202021}. \cite{https://link.springer.com/content/pdf/10.1007/s10869-011-9232-7.pdf} found that the order of communication medium influenced perceived inclusion, which in turn influenced individual participation. The study suggests that within diverse groups in which minority member is different from the rest of team and may feel excluded, initial computer-mediated communication may help the minority member feel more included. Furthermore, an online format is more environmentally-friendly since it eliminates travel-related emissions \cite{sarabipourChangingScientificMeetings2021,ninerBetterWhomLeveling2021, gattrellComparisonCarbonCosts2021}.

%Yanina: I add the reference about people feeling better using on-line and computer mediated communication when belong to a excluded group.

Alternatively, a conference could have a hybrid format with an in-person and online component. This dual format could allow a group of people to interact face-to-face while providing many others the opportunity to participate remotely. The challenge and requirement for this kind of setting would be to make the online component as relevant as the in-person component and not just a consolation prize to the less privileged in the community \cite{ninerBetterWhomLeveling2021}. %Adithi: After the conference is over, the conference materials and videos can be uploaded on free platforms, so that people who were left out of the main event due to any reason, still feel they were cared for and can go to the resources. Also when the event is online, a more timezone friendly approach could be used, all the events if online can be global friendly, people with dependents need hours to take care and the night to rest, if the conference is overnight, they lose sleep and suffer, instead if the conference could be friendly for all regions on certain days just like useR! 2021. 
% Rocío: we need to add that just because there's an online component, it doesn't mean that the in-person part of hybrid conferences can be exclusionary.

Take advantage of the online setting for new networking activities that can be inclusive for people with different socialization styles and accessibility needs. Social events in in-person conferences are often the most exclusive (i.e., galas, dinner nights at expensive venues), and they are a very common place where you have code of conduct violations.
% Rocío: I think that we should justify why we say that "social events in in-person conferences are often the most exclusive"
%ast: exclusionary? i added a couple of examples

Having them directly translated into their virtual versions does not address the need for creating opportunities for people to meet and bond. An online event is a great opportunity to create spaces for people who do not enjoy in person networking. Some ideas in this line are: offering the option of written chat only, instead of voice or video conversations, opening events with team work like trivia, offer some events that can be passively enjoyed like movies, yoga sessions, or art displays, where attendants can choose to just sit and enjoy without talking, or have a chat channel to comment on their experiences during the session. Respect people’s limits, preferences, and remember that “the usual” does not necessarily work with your community, and that one single networking activity will never serve the whole community. It is worth trying new and varied activities that might work with subgroups of attendees.

If the conference is online and synchronic (e.g. with live sessions), timezones should be considered: you may change the favored timezone year to year, or work on several timezones in each conference edition. If the conference is hybrid or in-person, be sure the location does not favor the same people over years, move from one continent to another.

@@

>>>>

The rules written here are directed to people who are part of a stable meetings committee that oversees the site/location selection process, that coordinates with the local organizers of conferences, or even the local/virtual organizers who desire to make an inclusive conference starting at the planning stage.
These tips stem from the authors' experience of organizing useR! 2021, a virtual and global statistical computing conference for users and developers of the R programming language \cite{r_core_team_2021}. 
We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, truly inclusive and accessible experience for everyone. 
Here, we share the lessons learned within the past year of organizing this global useR!2021, summarized as 10 simple rules towards an inclusive conference.
% Rocío: Could add a line introducing Fig. in https://docs.google.com/drawings/d/1iS1pLc9OldLMe_jhNT6OGIbT9s5KvlHpUoDd19vHkCY/edit?usp=sharing

@@

%This is specially important for networking spaces, where receiving ideas, direct feedback, and spaces to co-lead from a wide variety of community members can be the key to organize meaningful sessions for all the conferences attendees. This gave a sense of variety and an appeal of “at least one networking session for everyone”.  (Marcela)
%ast 17sep: i think we can talk here about planning neworking here, but also: the whole paragraph applies to team dynamics in general. Sara: maybe you can use this as we spoke relating team work here. And I think the working together part is not here but the following paragraph 
%sara: The last sentence in the paragraph sounds odd to me, but maybe I'm missing something. Andrea, I added a few sentences to the following paragraph regarding work dynamics
% Rocío: the rule is getting too long and, since this is focused on networking, I'd prefer it to go to a networking section.

@@

Make sure that every selection committee--the committee looking for keynote speakers, the selection committee for abstracts, the prizes and award committees--are also diverse \cite{swartzScienceValueDiversity2019, wongBuildDiversityScience2020}, and ask them to be aware that everyone has implicit biases, to recognize them, and try to counteract them. 
An inclusive and diverse organizing team (\textbf{Rule \ref{rule_organizing_team}}) is already a great starting point to overcome this bias in other roles. The regional and local communities in your field are also good sources to tap into. 
For example, for useR!2021, groups like AfricaR (africa-r.org), R-Ladies (rladies.org), MiR (mircommunity.com), Forwards (forwards.github.io), and LatinR (latin-r.com) were fundamental to reach people for the organizing team, potential presenters, to co-lead social events, and attendees; some of them even held spaces to help the members of their groups to prepare abstracts for submission.
When organizing an international academic conference, you may reach local associations, student organizations, or other Early Career Researchers groups or communities of practice: ask for their requirements and barriers, and explicitly invite them to contribute to the conference. % Rocío: ask about their requirements and barriers sounds a bit confusing.
% Rocío: Not everyone knows what a community of practice is. Could we add a definition or use other words? Definition: "Communities of practice are groups of people who share a concern, a set of problems, or a passion about a topic, and who deepen their knowledge and expertise in this area by interacting on an ongoing basis". (Wenger, E., McDermott, R. A., & Snyder, W. (2002). Cultivating Communities of Practice: A Guide to Managing Knowledge. Harvard Business Press.)
Since many conferences are held with the primary goal of displaying cutting-edge work led by senior scientists and academics, early career researchers are often disadvantaged. Diversity should spread among all speaking roles, and the conference should also help raise the profiles of early career researchers, fostering collaboration and building their skills.

Furthermore, consider bringing to the spotlight the people that have contributed to your field in more collaborative ways \cite{cheng2020x+}. You can, for instance, reframe the awards ceremony to acknowledge those who prepared accessible slides and presentations (see Rule \ref{rule_accessibility}), to acknowledge being mindful of inclusiveness. Community building, for instance, is challenging and usually unrewarded when compared to publications or software development \cite{acionWhyChooseCommunity2020}; give the people who deliberately took the time to work on their communities the recognition they deserve. Defy the stereotypical criteria for success by acknowledging these community practices. 
Finally, do not restrict people from marginalized groups or community-builders to talk or work only in issues related to Diversity, Inclusion, and Accessibility. Recognize their areas of expertise and respect their will regarding participation (or not) in community-building events.
% Rocío: Since the system usually rewards the most competitive people, appealing cvs and so on, I think it's worth it to explain that rewarding collaborative work is also part of unbiasing; not only because it's mostly done by women, people of color, etc., but because this kind of amazing work is usually neglected in our conferences. It may not be obvious to everyone that the content of this paragraph is not only "bring these people to the spotlight", but that it also means unbiasing the conference. I don't think I'll have time to add lines about this in the beginning of the paragraph so if anybody could do that, it would be great.

@@

In-person interaction in conferences is priceless, but only for the ones who can afford to attend. 
Barriers such as cost of registration, transport and accommodation, the logistics of long-distance travel, and discriminatory visa applications, are particularly true for global conferences that usually take place in high-income countries \cite{arendDisparityConferenceRegistration2019,gewinWhatScientistsShould2019}. 
% Rocío: we can add our Nature Correspondence letter to the citations later on. 
Online conferences are more inclusive: they do not require a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities.
% Rocío: cite Saliba 2020
This means that online conferences have a greater reach, not only in terms of participants but in terms of the tutors and presenters that can participate \cite{atkinsonJournalMedicine20202021}.
% Rocío: we can add Roosetal. and our Nature Correspondence letter to the citations here too. I would've liked one that talks about increase in participation of tutors and presenters but haven't found any. 
The online format may also make it easier to be inclusive of geographic regions by encompassing several timezones in the week or easily deciding to rotate the favored timezone year to year, without depending on a conference central location. 
% Rocío: main idea of the paragraph: online conferences are more inclusive

Networking and socializing in online conferences may be challenging, mostly because we are used to personal interactions at coffee or lunch breaks, or other in-person social settings. %Rocío: cite Saliba 2020. 
However, social events in in-person conferences can be exclusive (e.g. galas or dinner nights at expensive venues), and they are a common place for code of conduct violations to occur. % Rocío: any reference to back the last statement? The CoC guide?
Organizers should invest time in creating opportunities to meet and bond virtually. 
An online event is a great opportunity to create spaces for people who do not enjoy in person networking. 
Some ideas in this line are: offering the option of written chat only, instead of voice or video conversations, opening events with team work like trivia, offer some events that can be passively enjoyed like movies, yoga sessions, or art displays, where attendants can choose to just sit and enjoy without talking, or have a chat channel to comment on their experiences during the session. Respect people's limits, preferences, and remember that `the usual' does not necessarily work with everyone in your community, and that one single networking activity will never serve the whole community. It is worth trying new and varied activities that might work with subgroups of attendees.
Moreover, virtual communication may make people from minoritized groups feel more included, thus participate more (e.g. CITATION). % RJ: https://link.springer.com/content/pdf/10.1007/s10869-011-9232-7.pdf
% Rocío: Networking paragraph.

In addition to being more inclusive, an online format is more environmentally-friendly since it eliminates travel-related emissions \cite{sarabipourChangingScientificMeetings2021,ninerBetterWhomLeveling2021, gattrellComparisonCarbonCosts2021}. 
Besides, if conference materials are systematically stored and pre-recorded, they can be made available to people who missed them live or even after the conference is over. 
% Rocío: Short paragraph of additional advantages of online conferences that are not about inclusion.

Alternatively, a conference could have a hybrid format with an in-person and online component. This dual format could allow a group of people to interact face-to-face while providing many others the opportunity to participate remotely. The challenge and requirement for this kind of setting would be to make the online component as relevant as the in-person component and not just a consolation prize to the less privileged in the community \cite{ninerBetterWhomLeveling2021}.
Needless to say, an inclusive online component of a hybrid conference is not a free pass to have an exclusionary in-person component. % Rocío: I know this is a bit aggressive so I wouldn't mind if anybody helps with the wording. 
For instance, to avoid favoring the same regions in every edition, we suggest rotating the location of the conference—if it is global, moving from one continent to another. 
% Rocío: Hybrid conferences paragraph.

@@

% Rocío: This ref (https://www.nature.com/articles/d41586-020-01489-0 already in the .bib) can be cited for networking
